Multi-target detection with application to cryo-electron microscopy

### 2019-03-01 inverse_problems/inverse_problems.tex
./

%Our approach is based on considering the first few entries of the observation's autocorrelations. These entries correspond to products of $y$ with its shifted version, for some small shifts.
%If $s$ would be too clustered, then

@@

\TODO{Careful that $\gamma$ changed by a factor of $L$: need to modify here too.} Specifically, for some parameter $\gamma > 0$, the total number of signal occurrences is drawn from a Poisson distribution with expectation $(N-L+1)\gamma$. For each occurrence, a starting location $i$ is drawn uniformly and independently at random from $\{0, \ldots, N-L\}$. Then, one of the signals $x_1, \ldots, x_K$ is selected independently at random from a fixed distribution and added to the observation $y$ with first entry positioned at $i$. 
By definition of the Poisson process, the entries of $s$ are independent and follow a Poisson distribution with expectation $\gamma$. In particular, this model allows for (additive) signal overlap. In Section~\TODO{ref?}, we show that the autocorrelations of $y$ are nonetheless equivalent to those arising under our well-separated model above.

@@

In what follows, $x$ is the random variable corresponding to signal occurrences in $y$. In particular, for the model~\eqref{eq:model}, 
%$x=x_k$ with probability $\pi_k$.
%In particular, for
$x$ is sampled from $\{ x_1, \ldots, x_K \}$ with probabilities $(\pi_1, \ldots, \pi_K)$, so that its autocorrelations are given in explicit form as:
\begin{align}
	a_x^q  & = \sum_{k=1}^{K} \pi_ka_{x_k}^q.
	\label{eq:mixedautocorr}
\end{align}
We are given one observation (one realization) of $y$. Thus, we cannot compute the autocorrelations of $y$ exactly as they involve taking an expectation against the distribution of $y$. However, by the law of large numbers, as $N$ grows to infinity with $\gamma$ remaining constant, the empirical autocorrelations of $y$ converge to the actual autocorrelations of $y$, that is, % (ergodicity)
\begin{align}
\lim_{N \to \infty, \gamma \textrm{ constant}} \frac{1}{N} \sum_{i=-\infty}^{\infty} y[i]y[i+\ell_1]\cdots y[i+\ell_{q-1}] = a_y^q[\ell_1, \ldots, \ell_{q-1}].
\end{align}
This provides a concrete means of estimating the quantities $a_y^ q$. In the remainder of this section, we relate the observables $a_y^q$  to the unknowns $a_x^q$, first under the well-separated model, then under the Poisson model.
\TODO{Explicit expressions of the autocorrelations for cryo-EM are given in~\cite{bendory2018toward}}

@@

We will show the following expressions for the autocorrelations of $y$, expressed in terms of the autocorrelations of $x$ and the parameters $\gamma$ and $\sigma$:
%
\begin{align}
a_y^1  &=  \gamma a_{x}^1, 
\label{eq:mean_micrograph2} \\
a_y^2[\ell] &=  (\gamma a_x^1)^2 + \gamma a_x^2[\ell] + \sigma^2 \delta[\ell], 
\label{eq:ac2_micrograph2} \\
a_y^3[\ell_1,\ell_2]  & = (\gamma a_x^1)^3 + \gamma a_x^1  
\cdot ( \gamma a_x^2[\ell_1] + \gamma a_x^2[\ell_2]
+ \gamma a_x^2[\ell_2-\ell_1]) + \gamma a_x^3[\ell_1,\ell_2] 
\nonumber \\
& + \gamma a_x^1 \sigma^2 \big(\delta[\ell_1]+\delta[\ell_2] +\delta[\ell_1-\ell_2] \big).
\label{eq:ac3_micrograph2}
\end{align}

From these values, one may solve for the terms $\gamma a_x^1$, $\gamma a_x^2[\ell] + \sigma^2 \delta[\ell]$, and $\gamma a_x^3[\ell_1,\ell_2]$. Indeed, recovering  $\gamma a_x^1$ and $\gamma a_x^2[\ell] + \sigma^2 \delta[\ell]$ is immediate from \eqref{eq:mean_micrograph2} and \eqref{eq:ac2_micrograph2}, and $\gamma a_x^3[\ell_1,\ell_2]$ then follows from
%
\begin{align}
\gamma a_x^3[\ell_1,\ell_2]
= a_y^3[\ell_1,\ell_2] - (\gamma a_x^1)^3
- \gamma a_x^1 \cdot \big(a_y^2[\ell_1]  
+ a_y^2[\ell_2] + a_y^2[\ell_1-\ell_2]- 3(\gamma a_x^1)^2 \big).
\end{align}

\TODO{Does the equivalence hold for $\ell\neq 0$?}
\begin{corollary}
	The autocorrelations are equivalent...
\end{corollary}

It is now easy to show how to solve for both $\gamma$ and $\sigma$ from these expressions, assuming $x$ is generic. Indeed, we observe the product
%
\begin{align}
(\gamma a_x^1) (\gamma a_x^2[1])
&=  \frac{\gamma^2}{L^2} \left( \sum_{i=0}^{L-1} x[i] \right)
\left(\sum_{j=0}^{L-2} x[j]x[j+1] \right)
\nonumber \\
&= \frac{\gamma^2}{L^2} \sum_{j=0}^{L-2} \sum_{i=0}^{L-1} x[i]x[j]x[j+1]
\nonumber \\
&= \frac{\gamma^2}{L} \left\{ \sum_{\ell=0}^{L-1} a_x^3[1,\ell]
+ \sum_{\ell=1}^{L-2} a_x^3[\ell,\ell+1]  \right\}.
\end{align}
%
Since we also observe $\sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell] + \sum_{\ell=1}^{L-2} \gamma a_x^3[\ell,\ell+1]$, \TODO{Is there a bias?} we can form the ratio and solve for $\gamma$:

\begin{comment}
In this section we omit the affect of the noise as it remains the same for the well-separated model. We aim to show that the moments of

@@

\begin{proposition} \label{prop:gamma}
	\TODO{Can we do this simultaneously for Poisson? or do we only need to do it for well-separated and invoke the equivalence claim? Not if that claim requires knowledge of $\gamma, \sigma$.}
	Let $\sigma > 0$ be fixed and assume that the separation condition~\eqref{eq:spacing} holds. If the mean of $x$ is nonzero, then 
	\begin{equation}
	\gamma  \aseq \lim_{N \to \infty}\frac{L (a^1_y)^2}{a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\sigma^2}.
	\end{equation}
\end{proposition}

@@

@@

\TODO{Check and specify if this holds for both models}
\TODO{Did we define densities $\gamma_k$?}

@@

Autocorrelations in $d$ dimensions is defined for  $\ell_1,\ldots,\ell_{q-1}\in\mathbb{Z}^d$  as 
\begin{equation}
a_z^q[\ell_1,\ldots,\ell_{q-1}]   = \E\left\{\frac{1}{m^d} \sum_{i\in\mathbb{Z}^d} z[i]z[i+\ell_1]\cdots z[i+\ell_{q-1}]\right\}.
\label{eq:ac_d_dimension}
\end{equation}
Interestingly, for dimensions greater than one almost all deterministic (that is, the distribution of $z$ is a Delta function) \TODO{Do we call it the homogeneous case?} signals are determined uniquely from their second-order autocorrelation, up to two symmetries: sign (or phase for complex signals) and reflection through the origin (with conjugation in the complex case)~\cite{hayes1982reconstruction}. 
If the mean of signal is available and non-zero, the sign symmetry can be resolved. However, determining the reflection symmetry still requires additional information, beyond the second-order autocorrelation.
The case of 1-D signals is essentially different: generally there are $2^{L-2}$ signals with the same second-order autocorrelation (after eliminating  symmetries)~\cite{beinert2015ambiguities,bendory2017fourier}.

>>>>

@@

\TODO{Careful that $\gamma$ changed by a factor of $L$: need to modify here too.} Specifically, for some parameter $\gamma > 0$, the total number of signal occurrences is drawn from a Poisson distribution with expectation $(N-L+1)\gamma$. For each occurrence, a starting location $i$ is drawn uniformly and independently at random from $\{0, \ldots, N-L\}$. Then, one of the signals $x_1, \ldots, x_K$ is selected independently at random from a fixed distribution and added to the observation $y$ with first entry positioned at $i$. 
By definition of the Poisson process, the entries of $s$ are independent and follow a Poisson distribution with expectation $\gamma$. In particular, this model allows for (additive) signal overlap. %In Section~\TODO{ref?}, we show that the autocorrelations of $y$ are nonetheless equivalent to those arising under our well-separated model above.

@@

In what follows, $x$ is the random variable corresponding to signal occurrences in $y$. In particular, for the model~\eqref{eq:model}, 
%$x=x_k$ with probability $\pi_k$.
%In particular, for
$x$ is sampled from $\{ x_1, \ldots, x_K \}$ with probabilities $(\pi_1, \ldots, \pi_K)$, so that its autocorrelations are given in explicit form as:
\begin{align}
	a_x^q  & = \sum_{k=1}^{K} \pi_ka_{x_k}^q.
	\label{eq:mixedautocorr}
\end{align}
We are given one observation (one realization) of $y$. Thus, we cannot compute the autocorrelations of $y$ exactly as they involve taking an expectation against the distribution of $y$. However, by the law of large numbers, as $N$ grows to infinity with $\gamma$ remaining constant, the empirical autocorrelations of $y$ converge to the actual autocorrelations of $y$, that is, % (ergodicity)
\begin{align}
\lim_{N \to \infty, \gamma \textrm{ constant}} \frac{1}{N} \sum_{i=-\infty}^{\infty} y[i]y[i+\ell_1]\cdots y[i+\ell_{q-1}] = a_y^q[\ell_1, \ldots, \ell_{q-1}].
\end{align}
This provides a concrete means of estimating the quantities $a_y^ q$. In the remainder of this section, we relate the observables $a_y^q$  to the unknowns $a_x^q$, first under the well-separated model, then under the Poisson model.
\TODO{Explicit expressions of the autocorrelations for cryo-EM are given in~\cite{bendory2018toward}.}

@@

We will show the following expressions for the autocorrelations of $y$, expressed in terms of the autocorrelations of $x$ and the parameters $\gamma$ and $\sigma$:
%
\begin{align}
a_y^1  &=  \gamma a_{x}^1, 
\label{eq:mean_micrograph2} \\
a_y^2[\ell] &=  (\gamma a_x^1)^2 + \gamma a_x^2[\ell] + \sigma^2 \delta[\ell], 
\label{eq:ac2_micrograph2} \\
a_y^3[\ell_1,\ell_2]  & = (\gamma a_x^1)^3 + \gamma a_x^1  
\cdot ( \gamma a_x^2[\ell_1] + \gamma a_x^2[\ell_2]
+ \gamma a_x^2[\ell_2-\ell_1]) + \gamma a_x^3[\ell_1,\ell_2] 
\nonumber \\
& \qquad + \gamma a_x^1 \sigma^2 \big(\delta[\ell_1]+\delta[\ell_2] +\delta[\ell_1-\ell_2] \big).
\label{eq:ac3_micrograph2}
\end{align}

%
%From these values, one may solve for the terms $\gamma a_x^1$, $\gamma a_x^2[\ell] + \sigma^2 \delta[\ell]$, and $\gamma a_x^3[\ell_1,\ell_2]$. Indeed, recovering  $\gamma a_x^1$ and $\gamma a_x^2[\ell] + \sigma^2 \delta[\ell]$ is immediate from \eqref{eq:mean_micrograph2} and \eqref{eq:ac2_micrograph2}, and $\gamma a_x^3[\ell_1,\ell_2]$ then follows from
%%
%\begin{align}
%\gamma a_x^3[\ell_1,\ell_2]
%= a_y^3[\ell_1,\ell_2] - (\gamma a_x^1)^3
%- \gamma a_x^1 \cdot \big(a_y^2[\ell_1]  
%+ a_y^2[\ell_2] + a_y^2[\ell_1-\ell_2]- 3(\gamma a_x^1)^2 \big).
%\end{align}

%\begin{corollary}[informal for now] \label{cor:ac_equivalence}
%The first and the second autocorrelations  in the well-separated and the Poisson models are equivalent. 
%If $\sigma$ is known, then it is also true for the third-order autocorrelation. If $\sigma$ is unknown, the it is true whenever $\ell_1\neq 0$, $\ell_2\neq 0$ and $\ell_1\neq \ell_2$. 
%\end{corollary}
%
%It is now easy to show how to solve for both $\gamma$ and $\sigma$ from these expressions, assuming $x$ is generic. Indeed, we observe the product
%%
%\begin{align}
%(\gamma a_x^1) (\gamma a_x^2[1])
%&=  \frac{\gamma^2}{L^2} \left( \sum_{i=0}^{L-1} x[i] \right)
%\left(\sum_{j=0}^{L-2} x[j]x[j+1] \right)
%\nonumber \\
%&= \frac{\gamma^2}{L^2} \sum_{j=0}^{L-2} \sum_{i=0}^{L-1} x[i]x[j]x[j+1]
%\nonumber \\
%&= \frac{\gamma^2}{L} \left\{ \sum_{\ell=0}^{L-1} a_x^3[1,\ell]
%+ \sum_{\ell=1}^{L-2} a_x^3[\ell,\ell+1]  \right\}.
%\end{align}
%%
%Since we also observe $\sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell] + \sum_{\ell=1}^{L-2} \gamma a_x^3[\ell,\ell+1]$, \TODO{Is there a bias?} we can form the ratio and solve for $\gamma$: 
%
%\begin{align}
%\gamma = L \frac{(\gamma a_x^1) (\gamma a_x^2[1])}
%{\sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell] 
%	+ \sum_{\ell=1}^{L-2} \gamma a_x^3[\ell,\ell+1]}.
%\end{align}
%%With $\gamma$ in hand, we can estimate $\sigma^2$:
%%
%\begin{align}
%\sigma^2 = a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\frac{L (a^1_y)^2}{\gamma}
%- (2 L - 1) (a_y^1)^2.
%\end{align}
%
\begin{comment}
In this section we omit the affect of the noise as it remains the same for the well-separated model. We aim to show that the moments of

@@

\begin{proposition} \label{prop:gamma}
	Let $\sigma > 0$ be fixed and assume that the separation condition~\eqref{eq:spacing} holds. If the mean of $x$ is nonzero, then  \TODO{Should remove the limit}
	\begin{equation}
	\gamma  \aseq \lim_{N \to \infty}\frac{L (a^1_y)^2}{a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\sigma^2}.
	\end{equation}
\end{proposition}

@@

For the Poisson model: 
From these values, one may solve for the terms $\gamma a_x^1$, $\gamma a_x^2[\ell] + \sigma^2 \delta[\ell]$, and $\gamma a_x^3[\ell_1,\ell_2]$. Indeed, recovering  $\gamma a_x^1$ and $\gamma a_x^2[\ell] + \sigma^2 \delta[\ell]$ is immediate from \eqref{eq:mean_micrograph2} and \eqref{eq:ac2_micrograph2}, and $\gamma a_x^3[\ell_1,\ell_2]$ then follows from
%
\begin{align}
\gamma a_x^3[\ell_1,\ell_2]
= a_y^3[\ell_1,\ell_2] - (a_y^1)^3
- a_y^1 \cdot \big(a_y^2[\ell_1]  
+ a_y^2[\ell_2] + a_y^2[\ell_1-\ell_2]- 3(a_y^1)^2 \big).
\end{align}

\begin{proposition}
	Poisson result 
\end{proposition}
\begin{proof}
It is now easy to show how to solve for both $\gamma$ and $\sigma$ from these expressions, assuming $x$ is generic. Indeed, we observe the product
%
\begin{align}
(\gamma a_x^1) (\gamma a_x^2[1])
&=  \frac{\gamma^2}{L^2} \left( \sum_{i=0}^{L-1} x[i] \right)
\left(\sum_{j=0}^{L-2} x[j]x[j+1] \right)
\nonumber \\
&= \frac{\gamma^2}{L^2} \sum_{j=0}^{L-2} \sum_{i=0}^{L-1} x[i]x[j]x[j+1]
\nonumber \\
&= \frac{\gamma^2}{L} \left\{ \sum_{\ell=0}^{L-1} a_x^3[1,\ell]
+ \sum_{\ell=1}^{L-2} a_x^3[\ell,\ell+1]  \right\}.
\end{align}
%
Since we also observe $\sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell] + \sum_{\ell=1}^{L-2} \gamma a_x^3[\ell,\ell+1]$, \TODO{Is there a bias?} we can form the ratio and solve for $\gamma$:

\end{proof}

@@

\TODO{Check and specify if this holds for both models}
\TODO{Did we define densities $\gamma_k$?}
\TODO{We write the section for the Poisson model, but it would be the same for the Poisson model}

@@

Autocorrelations in $d$ dimensions is defined for  $\ell_1,\ldots,\ell_{q-1}\in\mathbb{Z}^d$  as 
\begin{equation}
a_z^q[\ell_1,\ldots,\ell_{q-1}]   = \E\left\{\frac{1}{m^d} \sum_{i\in\mathbb{Z}^d} z[i]z[i+\ell_1]\cdots z[i+\ell_{q-1}]\right\}.
\label{eq:ac_d_dimension}
\end{equation}
Interestingly, for the homogeneous case ($K=1$) in dimensions greater than one, almost all  signals are determined uniquely from their second-order autocorrelation, up to two symmetries: sign (or phase for complex signals) and reflection through the origin (with conjugation in the complex case)~\cite{hayes1982reconstruction}. 
If the mean of signal is available and non-zero, the sign symmetry can be resolved. However, determining the reflection symmetry still requires additional information, beyond the second-order autocorrelation.
The case of 1-D signals is essentially different: generally there are $2^{L-2}$ signals with the same second-order autocorrelation (after eliminating  symmetries)~\cite{beinert2015ambiguities,bendory2017fourier}.
